---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

## Loading and preprocessing the data

I use the various packages in the `tidyverse` (such as `dplyr`) for data 
wrangling, so I need to load the packages with:

```{r packages}
suppressPackageStartupMessages(library(tidyverse))
```

Then, we can simply import the data:

```{r import, cache=TRUE}
unzip("activity.zip")
activityData <- read_csv("activity.csv", col_types = "iDi")
```


## What is mean total number of steps taken per day?

Here, the activity data is summarized to produce an histogram of the total number
of steps taken each day.

```{r histogram, fig.height=4, fig.width=4.5}
totalSteps <- activityData %>% 
	group_by(date) %>% 
	summarize(steps = sum(steps))

hist(totalSteps$steps, col = "wheat",
     main = "Steps taken each day", xlab = "Total Steps")
```

Then, we calculate the mean and the median of the total number of steps.

```{r mean and median}
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
meanSteps
medianSteps <- median(totalSteps$steps, na.rm = TRUE)
medianSteps
```

## What is the average daily activity pattern?

To give an overview of the average daily activity, we calculate the average 
activity (**steps**) across all days, and plot it against the 5-minute interval 
*in each day*.  

We then calculate the moment in the day with the maximum average activity, and 
also add it to the graph with a red vertical line.

```{r daily activity}
averageActivity <- activityData %>% 
	group_by(interval) %>% 
	summarize(steps = mean(steps, na.rm = TRUE))

with(averageActivity,
     plot(interval, steps, type = "l", main = "Average daily activity",
          xlab = "5-minute interval in the day", ylab = "Average number of steps"))

maxValue <- averageActivity[averageActivity$steps == max(averageActivity$steps), ]
abline(v = maxValue$interval, col = "red")
maxValue$interval
```

## Imputing missing values

There are a number of days/intervals where there are missing values: we need 
to analyze the problem a bit in order to understand if it's necessary to input 
some of the missing value to have more reliable calculations and summaries.

Out of `r nrow(activityData)` total observations, the number of incomplete 
cases (i.e. with at least one missing value) is equal to:

```{r count NAs}
sum(!complete.cases(activityData))
```

This means that `r paste0(round(mean(!complete.cases(activityData))*100, 2), "%")` 
of the rows are incomplete.

To fill in the missing values, I decided to simply use the mean value for that 
5-minute interval. The mean is a **numeric** variable, but I decided to round 
the value to the nearest **integer**, to be consistent with the original data frame.

The following code inputs the missing values and creates a new data frame called
`completeData`:

```{r input NAs}
incompleteData <- activityData[!complete.cases(activityData), ]
filledData <- merge(incompleteData, averageActivity, by = "interval",
		    all.x = TRUE, all.y = FALSE)
filledData <- filledData %>% 
	mutate(steps = as.integer(steps.y)) %>%  # Rounding to the nearest integer
	select(steps, date, interval) %>% 
	tbl_df

completeData <- activityData[complete.cases(activityData), ]
completeData <- rbind(completeData, filledData) %>% 
	arrange(date, interval)
```

As requested, I now recalculate both the histogram, the mean and the median values.

```{r new histogram mean and median, fig.height=4, fig.width=9}
newTotalSteps <- completeData %>% 
	group_by(date) %>% 
	summarize(steps = sum(steps))

par(mfrow = c(1, 2))
hist(totalSteps$steps, col = "wheat",
     main = "Steps taken each day (with NAs)", xlab = "Total Steps")
hist(newTotalSteps$steps, col = "tan1",
     main = "Steps taken each day (without NAs)", xlab = "Total Steps")

newMeanSteps <- mean(newTotalSteps$steps, na.rm = TRUE)
newMedianSteps <- median(newTotalSteps$steps, na.rm = TRUE)
tab <- data.frame(With.NAs = c(meanSteps, medianSteps),
		  Without.NAs = c(newMeanSteps, newMedianSteps),
		  row.names = c("Mean", "Median"))
print(tab)
```

Filling in the missing values produced an alteration in the summaries: the new 
mean is lower than before (**`r round((newMeanSteps-meanSteps)/meanSteps*100, 2)`%** 
variation), and also the median is lower (
**`r round((newMedianSteps-medianSteps)/medianSteps*100, 2)`%** variation).

## Are there differences in activity patterns between weekdays and weekends?

Here I create a new variable to indicate whether a day is a **weekday** or
a **weekend** day.

```{r weekday and weekend}
Sys.setlocale("LC_TIME", "en_US.UTF-8") # Forcing English locale
library(forcats)
weekday <- as.factor(weekdays(completeData$date, abbreviate = TRUE))
completeData$day.type <- fct_recode(weekday,
				    "weekday" = "Mon", "weekday" = "Tue",
				    "weekday" = "Wed", "weekday" = "Thu",
				    "weekday" = "Fri",
				    "weekend" = "Sat", "weekend" = "Sun")
```

Then, I use the `ggplot2` package to create the panel plot:

```{r panel plot}
library(ggplot2)
by_daytype <- completeData %>% 
	group_by(day.type, interval) %>% 
	mutate(averageSteps = mean(steps))

g <- ggplot(by_daytype, aes(interval, averageSteps))
g + facet_grid(day.type ~ .) +
	geom_line() +
	labs(x = "5-minute interval in the day", y = "Average number of steps",
	     title = "Differences in average activity weekday vs. weekend")


```

Note that the average activity is different depending on the day type.